Supplementary Information

### 2021-08-31 doc/manuscript/ForestGEO_treerings_SI.Rmd
USGS editor comments
issue #115

year_trends_all_legend= "Figure S16. Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 species at 10 sites."

@@

traditional_comparison_generic_legend = "Shown are repsonses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis. Left panels show a month-by-month comparison of *beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable from species-level residual chronologies (traditional approach) and from individual-level analysis in *climwin* (approach presented here). Center panels compare the monthly *beta* coefficient estimates, with the dotted line indicating 1:1 correspondence. Finally, the right panels compare *beta* coefficients for the most influential window selected by *climwin*. Error bars indicate standard error of slope estimates. Note that 1:1 correspondence is not necessarily expected. See Appendix 5 for analysis methods and discussion of expected correspondence."

climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth. The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1). Difference in $AICc$ relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column). The third and fourth columns give values of linear and quadratic terms for each seasonal window. The final column shows the correlation of individual-level residuals to the selected climate driver, with the function fit in *climwin* (not the final fit). Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined."

GLS_generic_legend = " for all three growth metrics examined here. For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/composite_plots."

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
ss_table <- read.csv("tables_figures/sampling_details.csv", col.names = c("site", "species code", "n trees", "n cores", "n trees", "n cores", "sampled", "reconstructed*", "date range", "year effect analyzed"), stringsAsFactors = FALSE, check.names = FALSE)
kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'), longtable = T)%>%
  add_header_above(c(" ", " ", "all"=2, "with DBH"=2, "DBH range" = 2, " ")) %>%
  column_spec(2, width = "1cm") %>%
  column_spec(3:6, width = "1cm") %>%
  column_spec(10, width = "1.5cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position","repeat_header"), protect_latex = T) 
```
*Maximum reconstructed DBH's analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/dbh_to_bark_allometries_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(allometries_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
  column_spec(1, italic = T) %>%
  column_spec(6, width = "3cm") %>%
  column_spec(7, width = "6cm") %>%
  row_spec(0, italic = F, bold = T) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
For assignments of species as proxies for those with out available bark allometries, see Table `r species_table`.

---
nocite: | 
 @miles_specific_2009; @anderson-teixeira_size-related_2015; @zeibig-kichas_bark_2016
---

@@

\newpage
`r qualitative_results_comparison`, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Niobrara, Nebraska, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Little Tesuque, New Mexico, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Cedar Breaks, Utah, USA", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Scotty Creek, NW Territories, Canada", 10, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_speciesspecific_2021; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
---
*Indicates results from a regional study including but not limited to cores from the focal site.

@@

@@

`r year_validation_legend`. Panel (a) shows results the effect of year in the full GLS model (all significant with t-test’s p-value <0.05). Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Panels (b-e) show relationships between DBH and BAI by decade for each species individually. These plots visualize the data included in the GLS model, separated by decade, using a probability density function. Transparent ribbons indicate 95% confidence intervals. Parallel plots for all sites are available in the GitHub repository for this project: https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/supplementary_figures/decadal_plots.

@@

![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=85% }

@@

![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=85% }

@@

![`r GLS_SCBI_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=85% }

@@

![`r GLS_LT_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=85% }

@@

![`r SC_1970break_legend` *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }

@@

![`r RW_interactions_all_legend` Shown are modeled response functions for species at the minimum and maximum and maximum ends of the DBH range. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. Species are color coded as in Figure 6.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }

@@

\newpage
## SI References

>>>>

year_trends_all_legend= "Figure S16. Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 tree species at 10 sites."

@@

traditional_comparison_generic_legend = "Shown are responses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis. Left panels show a month-by-month comparison of *beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable from species-level residual chronologies (traditional approach) and from individual-level analysis in *climwin* (approach presented here). Center panels compare the monthly *beta* coefficient estimates, with the dotted line indicating 1:1 correspondence. Finally, the right panels compare *beta* coefficients for the most influential window selected by *climwin*. Error bars indicate standard error of slope estimates. Note that 1:1 correspondence is not necessarily expected. See Appendix 5 for analysis methods and discussion of expected correspondence."

climwin_generic = "Rows correspond to the four different models run: (1) all ring width (RW) data, (2) RW for the subset of cores for which diameter at breast height (DBH) could be reconstructed, (3) basal area increment (BAI), and (4) aboveground biomass (AGB) growth. The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1). Difference in in Akaike information criterion corrected for small sample size (AICc) relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column). The third and fourth columns give values of linear and quadratic terms for each seasonal window. The final column shows the correlation of individual-level residuals to the selected climate driver, with the function fit in *climwin* (not the final fit). Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined."

GLS_generic_legend = "For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/composite_plots."

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
ss_table <- read.csv("tables_figures/sampling_details.csv", col.names = c("site", "species code", "n trees", "n cores", "n trees", "n cores", "sampled", "reconstructed*", "date range", "year effect analyzed"), stringsAsFactors = FALSE, check.names = FALSE)
kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'), longtable = T)%>%
  add_header_above(c(" ", " ", "all"=2, "with DBH"=2, "DBH range" = 2, " ")) %>%
  column_spec(2, width = "1cm") %>%
  column_spec(3:6, width = "1cm") %>%
  column_spec(10, width = "1.5cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position","repeat_header"), protect_latex = T) 
```
*Maximum reconstructed diameters at breast height (DBH's) analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.

NA = not applicable (species was not included in analysis because of limited sample size).

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/dbh_to_bark_allometries_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(allometries_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
  column_spec(1, italic = T) %>%
  column_spec(6, width = "3cm") %>%
  column_spec(7, width = "6cm") %>%
  row_spec(0, italic = F, bold = T) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T)

# NOTE: references for this table are in the nocite list below table S5. Apparently refs in the first of two nocite sections don't appear in references list.
```
For assignments of species as proxies for those with out available bark allometries, see Table `r species_table`.

NA = not applicable (allometry based on published equation, value not reported).

@@

\newpage
`r qualitative_results_comparison`, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Niobrara, Nebraska, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Little Tesuque, New Mexico, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Cedar Breaks, Utah, USA", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Scotty Creek, NW Territories, Canada", 10, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
 @miles_specific_2009; @anderson-teixeira_sizerelated_2015; @zeibig-kichas_bark_2016
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_speciesspecific_2021; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
---
*Indicates results from a regional study including but not limited to cores from the focal site.

@@

PPT = precipitation, PET = potential evapotranspiration.

@@

`r year_validation_legend`. Panel (a) shows results the effect of year in the full GLS model (all significant with t-test’s p-value <0.05). Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Panels (b-e) show relationships between diameter breast height (DBH) and basal area increment (BAI) by decade for each species individually. These plots visualize the data included in the GLS model, separated by decade, using a probability density function. Transparent ribbons indicate 95% confidence intervals. Parallel plots for all sites are available in the GitHub repository for this project: https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/supplementary_figures/decadal_plots.

@@

![`r GLS_BCNM_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB). `r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/BCI.png){ width=85% }

@@

![`r GLS_SCBI_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB). `r GLS_generic_legend`  PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI.png){ width=85% }

@@

![`r GLS_SCBI_year_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB). `r GLS_generic_legend` PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=85% }

@@

![`r GLS_LT_year_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB). `r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=85% }

@@

![`r SC_1970break_legend` *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. PIMA = *Picea mariana*; number in parentheses in legend is sample size.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }

@@

![`r RW_interactions_all_legend` Shown are modeled response functions for species at the minimum and maximum and maximum ends of the DBH range. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. PDF = precipitation day frequency, PPT = precipitation, PET = potential evapotranspiration. Species are color coded as in Figure 6.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }

@@

*Note: Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.*

## SI References
